---
title: "Actuarial Severity Fitting for Limits and Deductibles"
author: "Greg Sollenberger"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Actuarial Severity Fitting for Limits and Deductibles}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(actuarialfitdist)
set.seed(20261001)
```

# What this package assumes

`actuarialfitdist` is designed around a claim-level **ground-up severity** likelihood.

This assumption is important enough to state repeatedly:

* `loss` is the ground-up loss severity.
* `loss` is not the insurer payment after deductible.
* `loss` is not an excess-of-deductible amount.
* `loss` is not an ACV/depreciated payment unless the analyst has already converted the experience to the intended common ground-up basis.
* The deductible column tells the fitter that otherwise-identical claims below the deductible are not observed.
* A finite limit tells the fitter that a claim recorded at or above the limit is right-censored there.

For an exact observation, the row likelihood is proportional to `f(loss) / S(deductible)`. For a limit-censored observation it is `S(limit) / S(deductible)`.

If a row has likelihood weight `w`, its log-likelihood contribution is `w` times the ordinary row log-likelihood. An integer weight of 2 therefore has exactly the same likelihood effect as repeating that claim row twice.

# 1. Auto property-damage limit analysis

Suppose an insurer is analyzing property-damage severity. There is no deductible in this example, but policies have different limits. A claim that reaches its policy limit is recorded at the limit and is right-censored.

```{r auto-limit-data}
n <- 1200
true_loss <- rlnorm(n, meanlog = log(9000), sdlog = 0.95)
policy_limit <- sample(c(10000, 25000, 50000, 100000), n, replace = TRUE,
                       prob = c(0.15, 0.35, 0.35, 0.15))

pd <- data.frame(
  loss = pmin(true_loss, policy_limit),
  deductible = 0,
  limit = policy_limit
)

mean(pd$loss >= pd$limit)
```

Fit several plausible severity families:

```{r auto-limit-fit}
pd_candidates <- fit_severity_candidates(
  pd,
  loss = "loss",
  limit = "limit",
  distributions = c("lognormal", "weibull", "gamma", "burr", "loglogistic")
)

pd_candidates$comparison
```

The package does not automatically choose one distribution. The statistical fit can be compared with the business output directly:

```{r auto-limit-factors}
pd_factors <- severity_factors(
  pd_candidates,
  deductible = 0,
  limit = c(10000, 25000, 50000, 100000, 250000),
  base_deductible = 0,
  base_limit = 25000
)

head(pd_factors)
```

The factor table is intentionally long-format. `compare_factors()` provides a side-by-side view when desired.

```{r auto-limit-side-by-side}
compare_factors(
  pd_candidates,
  metric = "rating_factor",
  deductible = 0,
  limit = c(10000, 25000, 50000, 100000),
  base_deductible = 0,
  base_limit = 25000
)
```

# 2. Auto collision deductible analysis

Now consider collision severity with varying deductibles and no finite policy limit. Ground-up losses below the deductible never appear in the claim sample.

```{r collision-data}
n <- 1500
collision_ground_up <- rgamma(n, shape = 2.2, scale = 3200)
collision_ded <- sample(c(250, 500, 1000, 2000), n, replace = TRUE,
                        prob = c(0.10, 0.35, 0.40, 0.15))
keep <- collision_ground_up > collision_ded

collision <- data.frame(
  loss = collision_ground_up[keep],
  deductible = collision_ded[keep],
  limit = Inf
)
```

The ground-up amount is retained. We do **not** subtract the deductible before fitting.

```{r collision-fit}
collision_fit <- fit_severity(
  collision,
  loss = "loss",
  deductible = "deductible",
  limit = "limit",
  distribution = "gamma"
)

summary(collision_fit)
```

Generate deductible factors relative to a $500 deductible:

```{r collision-factors}
severity_factors(
  collision_fit,
  deductible = c(250, 500, 1000, 2000, 5000),
  limit = Inf,
  base_deductible = 500,
  base_limit = Inf
)
```

# 3. Homeowners deductible analysis with varying Coverage A

For homeowners, claim severity can vary systematically with the insured amount. The optional `scale_by` column lets the fitted severity scale vary linearly with an analyst-supplied numeric variable.

Here Coverage A is the scaling variable. The same mechanism could instead use TIV, vehicle value, or a user-created transformation in another line of business.

```{r homeowners-data}
n <- 1600
coverage_a <- sample(seq(150000, 900000, by = 25000), n, replace = TRUE)
home_ded <- sample(c(1000, 2500, 5000, 10000), n, replace = TRUE)
shape <- 1.35
scale_coef <- 0.075
latent_scale <- scale_coef * coverage_a
home_ground_up <- rweibull(n, shape = shape, scale = latent_scale)
home_limit <- coverage_a
keep <- home_ground_up > home_ded

home <- data.frame(
  loss = pmin(home_ground_up[keep], home_limit[keep]),
  deductible = home_ded[keep],
  limit = home_limit[keep],
  coverage_a = coverage_a[keep]
)
```

Fit a Weibull model with zero intercept:

```{r homeowners-fit}
home_fit <- fit_severity(
  home,
  loss = "loss",
  deductible = "deductible",
  limit = "limit",
  distribution = "weibull",
  scale_by = "coverage_a",
  scale_intercept = FALSE
)

summary(home_fit)
```

When `scale_intercept = FALSE`, the intercept is exactly zero, not merely shrunk toward zero or omitted from printed output.

For the Weibull family, the native scale is `scale_coef * coverage_a`, so the fitted mean is that quantity times `Gamma(1 + 1 / shape)`. Other supported families use parameterizations that preserve the same intended proportional-mean relationship when their mean is finite.

Factor output can be produced for arbitrary Coverage A values:

```{r homeowners-factors}
severity_factors(
  home_fit,
  scale_value = c(200000, 400000, 600000, 800000),
  deductible = c(1000, 2500, 5000),
  limit = c(100000, 250000, 500000),
  base_deductible = 1000,
  base_limit = 250000
)
```

## Scaling diagnostics

A severity histogram across the entire Coverage A range is not a comparison to one fitted distribution: each Coverage A value implies a different fitted severity distribution. `fit_plot_data()` can therefore restrict the comparison to a meaningful scale range while still retaining the exact row-level scale, deductible, and limit values inside that range.

```{r homeowners-fit-plot, fig.width=7, fig.height=4}
plot_fit(
  home_fit,
  type = "density",
  scale_range = c(300000, 450000),
  nsim = 3,
  seed = 11
)
```

A separate scaling diagnostic asks whether the assumed relationship across Coverage A is reasonable.

```{r homeowners-scale-plot, fig.width=7, fig.height=4}
plot_scaling_fit(home_fit, bins = 10, nsim = 3, seed = 12)
```

The actual and modeled observed means are directly comparable because the modeled series reproduces each claim row's deductible and censoring structure. The fitted ground-up mean series is also shown to make the underlying severity-scaling relationship visible.

# 4. Spliced body-tail fitting

A single family may fit the body of a severity distribution reasonably while missing the upper tail. `fit_spliced_severity()` supports a body family below a threshold and a separate tail family above it.

The two densities are normalized on their respective sides of the threshold. A body probability is estimated. Density continuity is **not required**.

The threshold may be fixed by the analyst or searched over candidate values.

```{r splice-data}
n <- 1500
threshold_true <- 30000
is_body <- runif(n) < 0.90

# Draw body values from a lognormal conditional on being below the threshold.
Fb <- plnorm(threshold_true, log(9000), 0.75)
body_draw <- qlnorm(runif(n, 0, Fb), log(9000), 0.75)

# Draw Pareto-II tail values conditional on exceeding the threshold.
shape_tail <- 2.1
scale_tail <- 18000
Ft <- 1 - (1 + threshold_true / scale_tail)^(-shape_tail)
u_tail <- runif(n, Ft, 1)
tail_draw <- scale_tail * ((1 - u_tail)^(-1 / shape_tail) - 1)

splice_ground_up <- ifelse(is_body, body_draw, tail_draw)
splice_ded <- sample(c(0, 1000, 2500), n, replace = TRUE)
splice_lim <- sample(c(50000, 100000, 250000, Inf), n, replace = TRUE)
keep <- splice_ground_up > splice_ded

splice_dat <- data.frame(
  loss = pmin(splice_ground_up[keep], splice_lim[keep]),
  deductible = splice_ded[keep],
  limit = splice_lim[keep]
)
```

Search for a threshold:

```{r splice-fit}
splice_fit <- fit_spliced_severity(
  splice_dat,
  loss = "loss",
  deductible = "deductible",
  limit = "limit",
  body = "lognormal",
  tail = "pareto2",
  threshold_method = "search",
  threshold_probs = c(0.60, 0.95),
  threshold_grid = 4
)

splice_fit
splice_fit$threshold_search
```

The same factor-generation interface works for the spliced model:

```{r splice-factors}
severity_factors(
  splice_fit,
  deductible = c(0, 1000, 5000),
  limit = c(25000, 50000, 100000, 250000),
  base_deductible = 0,
  base_limit = 50000
)
```

# Valuation or depreciation at the payment stage

The fitting layer does not decide whether historical claims are RCV, ACV, or some other valuation basis. The analyst should first express the experience on a consistent ground-up basis.

At factor generation, a multiplicative valuation factor can then represent an ACV-type payment transformation. If the ground-up replacement-cost severity is `X`, the package evaluates payments of the form

```text
min(max(valuation_factor * X - deductible, 0), limit)
```

For example:

```{r valuation}
severity_factors(
  home_fit,
  scale_value = 400000,
  deductible = 2500,
  limit = 25000,
  valuation_factor = c(1.00, 0.80, 0.60, 0.40, 0.20),
  base_deductible = 2500,
  base_limit = 25000,
  base_valuation_factor = 1
)
```

# Weights and fixed parameters

A likelihood weight multiplies the row log-likelihood. Integer weights therefore behave as frequency weights.

```{r weights}
weighted_collision <- collision[1:1000, ]
weighted_collision$w <- sample(1:3, nrow(weighted_collision), replace = TRUE)

weighted_fit <- fit_severity(
  weighted_collision,
  loss = "loss",
  deductible = "deductible",
  limit = "limit",
  distribution = "gamma",
  weights = "w",
  fixed = list(shape = 2.2)
)

coef(weighted_fit)
```

# Hessian and bootstrap uncertainty

`vcov()` and `confint()` use a numerical Hessian of the optimized negative log-likelihood. These are large-sample Wald approximations; the Hessian is evaluated numerically rather than supplied as hand-derived second derivatives for every family.

```{r hessian}
vcov(collision_fit)
confint(collision_fit)
```

A lightweight parametric or case bootstrap is also available. Parametric bootstrap retains the fitted portfolio's deductible, limit, scaling, and weight structure.

```{r bootstrap, eval=FALSE}
boot <- bootstrap_fit(collision_fit, R = 5, type = "parametric", seed = 99)
head(boot$estimates)
boot$success_rate
```

For a searched spliced fit, `refit_threshold = FALSE` reuses the selected threshold in each bootstrap replicate by default. Set it to `TRUE` to repeat threshold search, understanding that this is much more computationally expensive.

# Custom ggplot2 and Plotly graphics

The plot-building data are part of the public interface:

```{r custom-data}
d <- fit_plot_data(home_fit, scale_range = c(300000, 450000), nsim = 3, seed = 5)
head(d)
```

This allows analysts to use their own graphics conventions without reproducing the statistically important observation logic. See `inst/examples/08_plotly_diagnostics.R` for Plotly examples.

# Optional terms, anomaly audits, and log-scale plots

The policy terms are optional. A sample with no deductible or censoring limit
can be fitted by naming only the loss column and family. In particular,
Coverage A is not a policy censoring limit merely because it is an insured value.

```{r new-api, fig.width=7, fig.height=4}
set.seed(20261008)
no_terms <- data.frame(loss = rlnorm(250, log(12000), .8))
fit_no_terms <- fit_severity(no_terms, loss = "loss",
  distribution = "lognormal", hessian = FALSE)
print(fit_no_terms)
plot_fit(fit_no_terms, x_scale = "log10", nsim = 2, seed = 11,
  xlim = c(1000, 100000))
```

If a supplied row is inconsistent with the observation mechanism, the default
is to exclude the row after warning; the original positional row number is
reported. This is an audit aid, not a substitute for understanding how the
original ground-up amounts were collected.

```{r excluded-rows}
problem <- no_terms
problem$deductible <- 0
problem$deductible[3] <- problem$loss[3] + 10
clean_fit <- suppressWarnings(fit_severity(problem, loss = "loss",
  deductible = "deductible", distribution = "lognormal", hessian = FALSE))
clean_fit$excluded_rows
clean_fit$excluded_details
```

For single-fit plots and multi-candidate plots, use `x_scale = "log10"` by
default; `xlim` zooms without discarding data during statistic calculation.
The histogram density is computed in log-severity coordinates when log-scaled.

```{r candidate-comparison-plot, fig.width=8, fig.height=4}
plot_candidate_fits(pd_candidates, type = "density", nsim = 2, seed = 11,
  x_scale = "log10", xlim = c(200, 150000))
```
